Synthesis of Linear Functional Programs

### 2021-07-16 article/article.tex

% TODO: Why linear types?
The \synname\ synthetiser combines the features of linear types with recursion,
parametric polymorphism, recursive polymorphic? algebraic data types, and
refinement types. The synthetiser is built on top of a system of proof-search in
linear logic. Proof-search relates to program synthesis through the Curry-Howard
correspondence~\cite{}, which states that \emph{propositions in a logic} map
directly to \emph{types}, and \emph{proofs} of those propositions to well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

The core of the synthesis is a \emph{sound} and \emph{complete} system
consisting of \emph{bottom-up} proof search in linear logic augmented with
resource management~\cite{} and focusing~\cite{} as techniques to, respectively,
solve resource related non-determinism, and eliminate from the proof-search
don't-care non-determinism. A more in depth explanation is in
section~\ref{sec:formal_system}.

With the core in place, we can synthetise programs from specifications that use
canonical linear types ($\lolli, \with, \tensor, 1, \oplus, \bang$), such as
fundamental linear logic equivalences. For example, given the specification $(A
\lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$, the program
\mintinline{haskell}{\a -> \b -> let c*d = b in a c d}
% $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$ $c$ $d$
(commonly known as \emph{uncurry}) is synthetised.

However, a programming language only with the canonical linear types lacks
expressiveness. To enhance \synname\ and its language, synthetise more
interesting programs -- and thus empirically prove the feasability of more
relevant synthesis in a linear context, we extend the syntax and type system
with recursiveness and richer types: -- polimorfic, refinements, adts
parametrizáveis...  todo

@@

All additions to the synthesis are extensions of the proof-search in linear
logic system -- in the end, the process still consists of constructing a proof.
Because we can only construct valid proofs (a wrong proof isn't a proof), we
know that all produced synthesis results are correct: \textbf{all programs are
correct by construction}, i.e. if we're able to synthetise it, then it typechecks,
and therefore fullfills the specification.

@@

@@

\bibliographystyle{splncs04}
\bibliography{references}
\end{document}

>>>>

% TODO: Why linear types?
The \synname\ synthetiser combines the features of linear types with recursion,
parametric polymorphism, recursive algebraic data types, and
refinement types. The synthetiser is built on top of a system of proof-search in
linear logic. Proof-search relates to program synthesis through the Curry-Howard
correspondence~\cite{}, which states that \emph{propositions in a logic} map
directly to \emph{types}, and \emph{proofs} of those propositions to well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

The core of the synthesis is a \emph{sound} and \emph{complete} system
consisting of \emph{bottom-up} proof search in propositional linear
logic based on \emph{focusing}~\cite{}. Our approach, being grounded
by propositions-as-types, ensures that all synthesized programs
(i.e.~proofs) are well-typed \emph{by construction} (i.e.~if the
synthesis procedure terminates, then the program satisfies its
specification). Moreover, we can leverage the modularity of the
proof-search based approach in two key axes: first, since proof search
need not construct only closed proofs, we can effectively synthesize
program sub-expressions (i.e.~synthesis based on typed holes);
secondly, the framework is amenable to extension beyond the core
propositional language, employing a richer type structure that
includes polymorphism, algebraic data types and type refinements.

\paragraph{A core linear calculus.}
we can synthetise programs from specifications that use
canonical linear types ($\lolli, \with, \tensor, 1, \oplus, \bang$), such as
fundamental linear logic equivalences. For example, given the specification $(A
\lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$, the program
\mintinline{haskell}{\a -> \b -> let c*d = b in a c d}
% $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$ $c$ $d$
(commonly known as \emph{uncurry}) is synthetised.

\paragraph{Beyond propositional logic.}
However, a programming language only with the canonical linear types lacks
expressiveness. To enhance \synname\ and its language, synthetise more
interesting programs -- and thus empirically prove the feasability of more
relevant synthesis in a linear context, we extend the syntax and type system
with recursiveness and richer types: -- polimorfic, refinements, adts
parametrizáveis...  todo

@@

@@

\bibliographystyle{splncs04}
\bibliography{references}

\appendix

@@

\end{document}
